Avoiding tipping points in fisheries management through Gaussian Process Dynamic Programming

### 2014-11-20 manuscripts/manuscript.Rmd
as submitted

<!-- Run R code for analysis, to be called by figures -->
```{r plot-options, message=FALSE, warning=FALSE, include=FALSE, echo=FALSE}
cbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", 
               "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

@@

```{r refresh-plot-options, message=FALSE, warning=FALSE, include=FALSE, echo=FALSE, cache=FALSE}
cbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", 
               "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
opts_chunk$set(dev='pdf', dev.args=c(version="1.7"))
opts_chunk$set(fig.width=4, fig.height=3, echo=FALSE)
theme_set(theme_bw(base_size=12))
```

@@

We develop a novel approach to address these concerns in the context of
fisheries; although the challenges and methods are germane to
other problems of conservation or natural resource exploitation.  Economic value and ecological concern have made
marine fisheries the crucible for much of the founding work on management
under uncertainty [@Gordon1954;
@Clark1976; -@Clark2009; @Reed1979; @May1979; @Ludwig1982]

<!-- Uncertainty outside the data without the correct model has not been handled. -->
<!-- __We don't have the model__ -->

@@

<!-- Point is: __We don't have the data where we need it__ -->
<!-- What do we call this?  Extrapolation uncertainty?  Pathological Uncertainty? -->

@@

Tipping points in ecological dynamics [@Scheffer2001;
@Polasky2011] highlight this problem because 
precise models are not available and data are
limited such as around  high
stock levels or an otherwise desirable state. With perfect information
, one would know just how far a system could
be pushed before crossing the tipping point, and management would be
simple.  But we face imperfect models and limited data and, with tipping points,  
even small errors can have
very large consequences, as we shall illustrate later.  Because intervention is often (but not always, see
@Hughes2013) too late after a tipping point has been crossed, management
is often concerned with avoiding 
tipping points before any data about them is available or the 
transition that would more clearly reveal a tipping point occurs
[e.g. @Bestelmeyer2012].

@@

<!--
 _Note on "not magic"_: honest uncertainty + SDP
The idea that any approach can perform well without either having to
know the model or have particularly good data should immediately draw
suspicion.  The reader must bear in mind that the strength of our approach
comes not from black-box predictive power from such limited information,
but rather, by providing a more honest expression of uncertainty outside
the observed data without sacrificing the predictive capacity near the
observed data. By coupling this more accurate description of what is known
and unknown to the decision-making under uncertainty framework provided
by stochastic dynamic programming, we are able to obtain more robust
management policies than with common parametric modeling approaches.
-->

@@

<!--

Although much argument can be made
over the 'correct' objective function, e.g. how to account for the
social value of fish left in the sea vs. the commercial value of fish
harvested; see @Halpern2013 for further discussion of this issue.  We can
always compare model performance across multiple potential objectives.
The important point is that a decision-maker does not necessarily need
a model that provides the best mechanistic understanding or the best
long-term outcome, but rather the one that delivers the best management
performance over the period of interest.
-->

@@

Dynamic optimization  requires characterizing the dynamics of a state
variable (or variables), a control action, and a value function.
For simplicity, we consider only a single state variable, which allows us to
focus on one-dimensional models of population dynamics 
for the comparison of parametric models and the GP.
This is a best-case scenario for the parametric models because we
simulate underlying dynamics from one of the three parametric models
, whereas in the natural world we never know the "true" mode
.  In addition, by choosing one-dimensional models with just a few parameters, we
limit the chance that poor performance will be due to inability to
estimate parameters accurately, something that becomes a more severe problem
for higher-dimensional parametric models. Finally, the parametric models
we consider are those most commonly used in modeling stock-recruitment
dynamics or to model sudden transitions between alternative stable states.

@@

<!-- 
- Statement of the models
-->

We let $X(t)$ denote the size (numbers or biomass) of the focal
population at time $t$ and assume that in the absence of take its
dynamics are.
.

\begin{equation}
X(t+1) = Z(t) f(X(t), p) \label{eq1} 
\end{equation}

Where Z(t) is log-normally distributed process stochasticity [@Reed1979]
and $p$ is a vector of parameters to be estimated from the data.
We describe the three choices for $f(X(t),p)$ in the next section.  
The control action is a harvest or take, $h(t)$, measured
in the same units as $X$, at time $t$.  Thus, in the presence of take,
the population size on the right hand side of Eqn 1 is replaced by
$S(t) = X(t) - h(t)$.

To construct the value function, we consider a return when $X(t) = x(t)$ and
harvest $h(t) = h$ denoted as the reward, $R(x(t), h)$. For example, if the
return is the harvest at time $t$, then $R(x(t), h(t)) = \min(x(t), h(t))$. We
assume that future harvests are discounted relative to current ones at
a constant rate of discount $\delta$ and ask for the harvest policy
that maximizes total discounted harvest between the current time $t$
and a final time $T$. That is, we seek to maximize over choices of
harvest $E_{X(t+1)} [ \sum_{t = 0}^{T}  R(X(t), h(t), t) \delta^t]$,
where the state dynamics are given by Eqn 1 and $E$ denotes the
expectation over the process stochasticity in the future population state.

@@

\begin{equation}
\mathbf{\mathrm{E}}_{X(t+1)} \left[ V( X(t+1),t+1) | x(t), h(t) \right] = \mathbf{\mathrm{E}}_{Z(t)} \left[V( Z(t) f(x(t) - h(t))|p), t+1 | x(t), h(t) \right]
\end{equation}

where the population size after the take is $x(t) - h(t)$, which is then
translated into $X(t+1)$ by Eqn 1 (that is, we replace $X(t+1)$ by $Z(t) f(x(t)-h(t)|p)$).

@@

\begin{equation}
\mathbf{\mathrm{E}}_{X(t+1)} \left[ V(X(t+1), t+1) | x(t), h(t) \right] = \mathbf{\mathrm{E}}_{p|\mathrm{data}} \{ \mathbf{\mathrm{E}}_{Z(t) | p, \mathrm{data}} \left[ V(Z(t) f(x(t) - h(t)|p), t)  \right] \}
\end{equation}

@@

\begin{equation}
\mathbf{\mathrm{E}}_{X(t+1)} \left[ V(X(t+1), t+1) | x(t), h(t) \right] = \mathbf{\mathrm{E}}_{p|\mathrm{data}} \{ \mathbf{\mathrm{E}}_{f, Z(t) | p, \mathrm{data}} \left[ V( Z(t) f(x(t) - h(t)|p), t) \right] \}
\end{equation}

@@

We consider a period of `r Tobs` in which data are obtained to estimate the parameters or the GP. This is long
enough that the estimates do not depend on the particular realization,
and longer times are not likely to provide substantial improvement.  
We fit each of the models is fit to the same data (Figure 1).

<!-- Bayesian inference of parametric models -->

@@

@@

The core difference for our purpose between the GP and the parametric
models is that the GP model is defined explicitly by the observed data.  That is, it cannot be specified by the value of parameters alone.

<!-- Statement of model -->

@@

A GP is a 
stochastic process for which any realization of for
a  finite sample of points $n$ is a multivariate normal distribution
of dimension $n$. Thus, to characterize the GP we need a mean function
and a covariance function. We proceed as follows.

As before, we assume that the data $X(t)$ are observed with process stochasticity around a mean function $ g(X(t))$

@@

where $\varepsilon$ are IID normal random variables with zero-mean and variance $\sigma_n^2$. Note that we have chosen to
assume additive stochasticity.  While we could consider log-normal stochasticity
as in the parametric models, we make this choice to emphasize that the
Gaussian process approach need not have structurally correct stochasticity to be effective.

Now consider the situation where we have a set of  previous observations of the _timeseries_, $X_{\textrm{obs}}(t)$,
and wish to _predict_ future values $X_{\textrm{p}}(t)$.  We consider the observations as pairs of points
$X_{\textrm{obs}}(t), X_{\textrm{obs}}(t+1)$ for $t=1, 2, …$ up through final observation time $T$.  As described above, in the  GP, a finite sample of these predictions at $n$ points into the future, $X_{\textrm p}(\tau)$ for $\tau \in [T+1, T+2, ..., T+n]$, is a multivariate normal ($\mathcal{N}$):

\begin{equation}
g(X_p|X_{\textrm{obs}}(t), X_{\textrm{obs}}(t+1)) \sim \mathcal{N}(E,C)
\end{equation}

with mean and covariance matrix given by (see Eqn 2.22 of @Rasmussen2006 for a more detailed derivation)

\begin{equation}
E = K(X_{\textrm p}(\tau), X_{\textrm{obs}}(t)) \left(K(X_{\textrm{obs}}(t),X_{\textrm{obs}}(t)) - \sigma \mathbb{I} \right)^{-1} X_{\textrm{obs}}(t+1)
\end{equation}

\begin{equation}
C = K(X_{\textrm p}(\tau), X_{\textrm p}(\tau)) - K(X_{\textrm p}(\tau), X_{\textrm{obs}}(t)) \left(K(X_{\textrm{obs}}(t),X_{\textrm{obs}}(t)) - \sigma \mathbb{I} \right)^{-1} K(X_{\textrm{obs}}(t), X_{\textrm p}(\tau))
\end{equation}

Where $K$, called the covariance kernel, measures how much influence one observation has on another.  In this paper, for $K$ use a matrix whose $(i,j)$th element is given by:

\begin{equation}
K_{i,j}(x_i, y_j) = \exp\left( \frac{ -(x_i - y_j)^2}{2 \ell^2} \right)
\end{equation}
so that $\ell$ gives the characteristic length-scale over which correlation between
two observations decays.

We use inverse Gamma priors on both the length-scale $\ell$ and $\sigma$, thus for example

@@

<!-- Inference of the BNP model -->

@@

To ensure our results are robust to the choice of parameters, we
consider 96 different scenarios, described in detail below.   To help  better
understand the process, we first describe in detail the
results of a single scenario.

```{r figure_1, dependson=c("assemble-models", "par-fns", "plot-options"), fig.cap="Points show the training data of stock-size over time.  Curves show the posterior step-ahead predictions based on each of the estimated models. Observe that all models are fitting the data reasonably well. In the GP panel, black lines are means of the GP for random samples from the posterior distributions, while grey bands around them reflect the standard deviations of the GP distribution.  (Recall that while given a particular set of parameters, a parametric model gives just one curve, the GP is always a distribution, whose mean we plot in black.)", fig.width=8, fig.height=6, cache=FALSE}
step_ahead_posteriors <- function(x){
  gp_f_at_obs <- gp_predict(gp, x, burnin=1e4, thin=300)
  df_post <- melt(lapply(sample(100, 30),  
  function(i){
    data.frame(time = 1:length(x), stock = x, 
                GP = mvrnorm(1, gp_f_at_obs$Ef_posterior[,i], gp_f_at_obs$Cf_posterior[[i]]),
                True = step_ahead(x,f,p),  
                MLE = step_ahead(x,f,est$p), 
                Allen = step_ahead(x, allen_f, pardist[i,]), 
                Ricker = step_ahead(x, ricker_f, ricker_pardist[i,]), 
                Myers = step_ahead(x, myers_f, myers_pardist[i,]))
  }), id=c("time", "stock"))
}
df_post <- step_ahead_posteriors(x)
figure1b_posteriors <- ggplot(df_post) + geom_point(aes(time, stock)) + 
  geom_line(aes(time, value, col=variable, group=interaction(L1,variable)), alpha=0.2) + 
  facet_wrap(~variable) + 
  scale_colour_manual(values=colorkey) +  
  theme(legend.position="none")
figure1b_posteriors

write.csv(df_post, "components/data/figure1.csv")
```

@@

<!-- 
### Figure 2 
- Data comes from limited region of state-space 
- (Should really show uncertainty of all models here.  Capture the forecast uncertainty several steps down the road?)   
-->

```{r figure_2, dependson=c("assemble-models", "plot-options"), fig.cap="The inferred Gaussian process compared to the true process and maximum-likelihood estimated process.  We show the expected value for the function $f$ under each model.  Two standard deviations from the estimated Gaussian process covariance with (light grey) and without (darker grey) measurement error are also shown.  The training data are also shown as black points.  The GP is conditioned on (0,0), shown as a pseudo-data point, as explained in the text", fig.width=8, fig.height=6, cache=FALSE}
x_grid_short <- x_grid[1:40]
gp_short <- gp_predict(gp, x_grid_short, burnin=1e4, thin=300)
models_posteriors <- 
  melt(lapply(sample(100, 50), 
              function(i){
    sample_gp <- mvrnorm(1, 
                            gp_short$Ef_posterior[,i],         
                            gp_short$Cf_posterior[[i]])
    data.frame(stock = x_grid_short, 
               GP = sample_gp,
               y = sample_gp,
               ymin = sample_gp - 2 * sqrt(gp_short$E_Vf), 
               ymax = sample_gp + 2 * sqrt(gp_short$E_Vf), 
               True = sapply(x_grid_short,f,0, p),  
               MLE = sapply(x_grid_short,f,0, est$p), 
               Allen = sapply(x_grid_short, allen_f, 0, pardist[i,]), 
               Ricker = sapply(x_grid_short, ricker_f, 0, ricker_pardist[i,]), 
               Myers = sapply(x_grid_short, myers_f, 0, myers_pardist[i,]))
             }), 
       id=c("stock", "y", "ymin", "ymax"))
ggplot(models_posteriors) + 
    geom_ribbon(aes(x=stock, y=y, ymin=ymin, ymax=ymax, group=L1), 
                  fill = "gray80", 
                  data=subset(models_posteriors, variable == "GP")) + 
    geom_line(aes(stock, value, col = variable, 
                  group=interaction(L1,variable)), alpha=0.2) + 
    geom_point(data = obs, aes(x,y)) + 
    xlab(expression(X[t])) + ylab(expression(X[t+1])) +
    facet_wrap(~variable) + 
    scale_colour_manual(values=colorkey) +  
    theme(legend.position="none")

@@

We show the  mean inferred population dynamics of each model relative to the
true model used to generate the data in Figure 2, predicting
the relationship between observed population size (x-axis) to the population
size after recruitment the following year.  
In addition to the raw data, the GP is conditioned
on going through the point 0,0 without error.  This conditioning on (0,0) is equivalent to making the assumption that the population is closed, so that once it hits 0 it stays at 0, despite the lack of any data in the observed sequence to justify this.  This assumption illustrates how the GP can capture common-sense biology without having to assume more explicit details about the dynamics at low population numbers that have never been observed. If the population were not closed, one could repeat the entire analysis without this assumption.  Unlike parametric models, the GP corresponds to a
distribution of curves - the gray band with mean shown in black. Uncertainty in the parameters of the GP (not shown) further widens the band of possible population sizes.
In Figure S1 (see supplement), we show the performance of the models outside the observed training data.

@@

```{r figure_4, dependson=c("sim"), fig.cap="In the management context, GPDP outperforms approaches based on parametric models. We show 100 replicate simulations of the stock dynamics (Eqn 1) under the policies derived from each of the estimated models, as well as the policy based on the exact underlying model.", fig.width=6, fig.height=6, cache=FALSE}
ggplot(sims_data) + 
  geom_line(aes(time, fishstock, group=interaction(reps,method), color=method), alpha=0.1) +
  scale_colour_manual(values=colorkey, guide = FALSE) +
  xlab("stock size, x(t)") +
  facet_wrap(~method) + 
  guides(legend.position="none", color=FALSE)

@@

Changing the intensity of the stochasticity or the distance between stable and unstable steady states
does not impact the performance of the GP relative to the optimal solution
obtained from the true model and true parameters.  The parametric models are
more sensitive to this difference.  Large values of $\sigma$ relative to the distance between the stable and unstable
point increases the chance of a stochastic transition.  More precisely,
if we let $L$ is the distance between the stable and unstable steady states, then the probability that fluctuations drive the population across the unstable steady state scales as
$\exp\left(\frac{L^2}{\sigma^2}\right)$

@@

```{r figure_6, fig.width=6, fig.height=3, dependson=c("sensitivity-trends"), fig.cap="The effect of increasing noise or decreasing Allee threshold levels on the net present value of the fishery when managed under the GPDP, relative to managing under the true model (with known parameters).  Other than the focal parameter (stochasticity, Allee threshold), other parameters are held fixed as above to illustrate this effect.", cache=FALSE}
plot_sigmas <- ggplot(vary_sigma, aes(noise, value)) + 
  geom_point() + ylab("Net present value") + 
  xlab("Level of growth stochasticity") + theme_bw() 
plot_allees <- ggplot(vary_allee, aes(allee, value)) + 
  geom_point()+ ylab("Net present value") + 
  xlab("Allee threshold stock size") + theme_bw()
multiplot(plot_sigmas, plot_allees, cols=2) 
write.csv(vary_sigma, "components/data/figure6a.csv")
write.csv(vary_allee, "components/data/figure6b.csv")
```

@@

Our results illustrate that model-choice approaches can be absolutely
misleading – by providing support to models that cannot capture tipping point
dynamics because they have fewer parameters and the data are far from the tipping point.  That is, when the data 
come from around the stable steady state,
all the parametric models are approximately linear and approximately
identical.  Thus, it is intuitive that all model selection methods choose the simplest model.
In a complex world, the result is suboptimal. But in a world that might
contain tipping points, the result is disastrous.

Many managers both in fisheries and beyond face a similar situation: they
have not observed the population dynamics at all possible densities.  A
lack of comprehensive data at all population sizes, combined with the
inability to formulate accurate population models for low population sizes
in the absence of data, makes this situation the rule more than the exception.
Relying on parametric models and model choice processes that favor simplicity
ignores this basic reality. For a long time,  Carl Walters [e.g. @Walters1978] has argued that if we began
by fishing any newly exploited population down to very low levels and
then let it recover, we would be much better at estimating population
dynamics and thus predicting the optimal harvest levels.  While certainly true,
this presents a rather risky policy in the face of potential tipping points. 
The GPDP offers an alternative way to acknowledge this uncertainty.

@@

<!--

### Risk-prone and risk-adverse value functions

The degree to which the decision-making part of the algorithm (the SDP)
chooses to explore or avoid the resulting region of uncertainty can also
be influenced by the curvature of the value (profit) function $R$. Both
to simplify the intuition and avoid biasing this result, we have chosen
profits that are linear in the catch and thus neither risk-prone nor
risk adverse.  Making this function concave, representing the typical
assumption of diminishing returns, would make the SDP more risk-adverse
(as larger-than-expected stock sizes offer diminished returns relative
to the cost of smaller-than-expected stock sizes), and strengthen the
result shown here in which the BNP solution tends to avoid the region
of uncertainty.  Sufficiently convex or risk-prone functions could lead
the SDP to attempt higher exploitation rates despite the uncertainty.
Understanding the relative roles of such functions would be a promising
direction for future investigation.

-->

>>>>

```{r plot-options, message=FALSE, warning=FALSE, include=FALSE, echo=FALSE}
cbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", 
               "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

@@

```{r refresh-plot-options, message=FALSE, warning=FALSE, include=FALSE, echo=FALSE, cache=FALSE}
cbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", 
               "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
opts_chunk$set(dev='pdf', dev.args=c(version="1.4"))
opts_chunk$set(fig.width=4, fig.height=3, echo=FALSE)
theme_set(theme_bw(base_size=12))
```

@@

We develop a novel approach to address these concerns in the context of
fisheries; although the challenges and methods are germane to
other problems of conservation or natural resource exploitation.  Economic value and ecological concern have made
marine fisheries the crucible for much of the founding work on management
under uncertainty [@Gordon1954;
@Clark1976; -@Clark2009; @Reed1979; @May1979; @Ludwig1982].

@@

@@

Tipping points in ecological dynamics [@Scheffer2001;
@Polasky2011] highlight this problem because 
precise models are not available and data are
limited such as around  high
stock levels or an otherwise desirable state. With perfect information
, one would know just how far a system could
be pushed before crossing the tipping point, and management would be
simple.  But we face imperfect models and limited data and, with tipping points,  
even small errors can have
very large consequences, as we shall illustrate later.  Because intervention may be too late once a 
tipping point has been crossed (but see @Hughes2013), management
is often concerned with avoiding 
tipping points before any data about them are available.

@@

@@

@@

Dynamic optimization  requires characterizing the dynamics of a state
variable (or variables), a control action, and a value function.
For simplicity, we consider only a single state variable, which allows us to
focus on one-dimensional models of population dynamics 
for the comparison of parametric models and the GP.
This is a best-case scenario for the parametric models because we
simulate underlying dynamics from one of the three parametric models
, whereas in the natural world we never know the "true" model.
In addition, by choosing one-dimensional models with just a few parameters, we
limit the chance that poor performance will be due to inability to
estimate parameters accurately, something that becomes a more severe problem
for higher-dimensional parametric models. Finally, the parametric models
we consider are commonly used in modeling stock-recruitment
dynamics or to model sudden transitions between alternative stable states.

@@

We let $X(t)$ denote the size (numbers or biomass) of the focal
population at time $t$ and assume that in the absence of take its
dynamics are:

\begin{equation}
X(t+1) = Z(t) f(X(t), \mathbf{p}) \label{eq1} 
\end{equation}

Where Z(t) is log-normally distributed process stochasticity [@Reed1979]
and $\mathbf{p}$ is a vector of parameters to be estimated from the data.
We describe the three choices for $f(X(t),\mathbf{p})$ in the next section.  
The control action is a harvest or take, $h(t)$, measured
in the same units as $X$, at time $t$.  Thus, in the presence of take,
the population size on the right hand side of Eqn 1 is replaced by
$S(t) = X(t) - h(t)$.

To construct the value function, we consider a return when $X(t) = x(t)$ and
harvest $h(t) = h$ denoted as the reward, $R(x(t), h)$. For example, if the
return is the harvest at time $t$, then $R(x(t), h(t)) = \min(x(t), h(t))$. We
assume that future harvests are discounted relative to current ones at
a constant rate of discount $\delta$ and ask for the harvest policy
that maximizes total discounted harvest between the current time $t$
and a final time $T$. That is, we seek to maximize over choices of
harvest $E [ \sum_{t = 0}^{T}  R(X(t), h(t), t) \delta^t]$,
where the state dynamics are given by Eqn 1 and $E$ denotes the
expectation over future population states.

@@

\begin{equation}
\mathbf{\mathrm{E}}_{X(t+1)} \left[ V( X(t+1),t+1) | x(t), h(t) \right] = \mathbf{\mathrm{E}}_{Z(t)} \left[V( Z(t) f(x(t) - h(t))|\mathbf{p}), t+1 | x(t), h(t) \right]
\end{equation}

where the population size after the take is $x(t) - h(t)$, which is then
translated into $X(t+1)$ by Eqn 1 (that is, we replace $X(t+1)$ by $Z(t) f(x(t)-h(t)|\mathbf{p})$).

@@

\begin{equation}
\mathbf{\mathrm{E}}_{X(t+1)} \left[ V(X(t+1), t+1) | x(t), h(t) \right] = \mathbf{\mathrm{E}}_{\mathbf{p}|\mathrm{data}} \{ \mathbf{\mathrm{E}}_{Z(t) | \mathbf{p}, \mathrm{data}} \left[ V(Z(t) f(x(t) - h(t)|\mathbf{p}), t)  \right] \}
\end{equation}

@@

\begin{equation}
\mathbf{\mathrm{E}}_{X(t+1)} \left[ V(X(t+1), t+1) | x(t), h(t) \right] = \mathbf{\mathrm{E}}_{\mathbf{p}|\mathrm{data}} \{ \mathbf{\mathrm{E}}_{f, Z(t) | \mathbf{p}, \mathrm{data}} \left[ V( Z(t) f(x(t) - h(t)| \mathbf{p}), t) \right] \}
\end{equation}

@@

We consider a period of `r Tobs` in which data are obtained to estimate
the parameters or the GP. This is long enough that the estimates do not
depend on the particular realization, and longer times are not likely
to provide substantial improvement.  Each of the models is fit to the
same data (Figure 1).

@@

Additionally, we compute the maximum likelihood estimate (MLE, as we will refer to 
this model in the figures) of the parameters for the (structurally correct) Allen model.
Comparing this to using the posterior distribution of parameters inferred from MCMC for
the same model gives some indication of the importance of this uncertainty in the 
dynamic programming.

@@

The core difference for our purpose between the estimated GP and the estimated parametric
models is that the estimated GP model is defined explicitly in reference to the observed data.
As a result, uncertainty arises in the GP model not only from uncertainty in the parameters,
but is also increases in regions farther from the observed states, such as low population
sizes in the example illustrated here.  The estimated parametric models, by contrast, are
completely specified by the parameters.

@@

A GP is a stochastic process for which any realization consisting of n
points follows a multivariate normal distribution of dimension $n$. To
characterize the GP we need a mean function and a covariance function. We
proceed as follows.

As before, we assume that the data $X(t)$ are observed with process
stochasticity around a mean function $g(X(t))$

@@

where $\varepsilon$ are IID normal random variables with zero-mean and variance $\sigma^2$. Note that we have chosen to
assume additive stochasticity.  While we could consider log-normal stochasticity
as in the parametric models, we make this choice to emphasize that the
Gaussian process approach need not have structurally correct stochasticity to be effective.

In order to make predictions, we update the GP based on the observed set of transitions.  To do so, we collect the time series of observed states into a vector of "current" states, $\mathbf{X}_{\textrm{obs}} = \{X(1), \dots, X(T-1)\}$ and a vector of "next" states $\mathbf{Y}_{\textrm{obs}} = \{X(2),\dots,X(T)\}$ where $T$ is the time of the final observation.  Conditional on these observations, the predicted next state, $g(X_p)$, for any given "current" state, $X_p$ follows a normal distribution with mean $E$ and variance $C$ determined using the standard rules for conditioning in multivariate normals, i.e.

\begin{equation}
E = K(\mathbf{X}_{\textrm p}, \mathbf{X}_{\textrm{obs}}) \left(K(\mathbf{X}_{\textrm{obs}},\mathbf{X}_{\textrm{obs}}) - \sigma \mathbb{I}_n \right)^{-1} \mathbf{Y}_{\textrm{obs}}
\end{equation}

and

\begin{equation}
C = K(\mathbf{X}_{\textrm p}, \mathbf{X}_{\textrm p}) - K(\mathbf{X}_{\textrm p}, \mathbf{X}_{\textrm{obs}}) \left(K(\mathbf{X}_{\textrm{obs}},\mathbf{X}_{\textrm{obs}}) - \sigma \mathbb{I} \right)^{-1} K(\mathbf{X}_{\textrm{obs}}, \mathbf{X}_{\textrm p})
\end{equation}

Here $\mathbb{I}_n is the $n$ by $n$ identity matrix (i.e. a matrix with ones down the diagonal and zeros elsewhere) and $K$ is the ‘covariance kernel.’  The covariance kernel controls how much influence one observation has on another.  In the present application we use the squared-exponential kernel which, when evaluated over a pair of vectors, say $x$ and $y$,  generates a covariance matrix whose $i,j$th element is given by

\begin{equation}
K_{i,j}(\mathbf{x}, \mathbf{y}) = \exp\left( \frac{ -(x_i - y_j)^2}{2 \ell^2} \right)
\end{equation}

so that $\ell$ gives the characteristic length-scale over which correlation between
two observations decays. See @Rasmussen2006 for other choices of covariance kernels and their properties.
Note that this simple formulation assumes a prior mean of zero. For the parameters we use inverse Gamma priors on both the length-scale $\ell$ and $\sigma$, thus for example

@@

@@

To ensure our results are robust to the choice of parameters, we
will consider 96 different scenarios.  To help better
understand the process, we first describe in detail the
results of a single scenario.

```{r figure_1, dependson=c("assemble-models", "par-fns", "plot-options"), fig.cap="Points show the training data of stock-size over time.  Curves show the expectations over the posterior step-ahead predictions based on each of the estimated models. Observe that all models are fitting the data reasonably well.)", fig.width=8, fig.height=6, cache=FALSE}
step_ahead_posteriors <- function(x){
  gp_f_at_obs <- gp_predict(gp, x, burnin=1e4, thin=300)
  df_post <- melt(lapply(sample(100, 30),  
  function(i){
    data.frame(time = 1:length(x), stock = x, 
                GP = mvrnorm(1, gp_f_at_obs$Ef_posterior[,i], gp_f_at_obs$Cf_posterior[[i]]),
                True = step_ahead(x,f,p),  
                MLE = step_ahead(x,f,est$p), 
                Allen = step_ahead(x, allen_f, pardist[i,]), 
                Ricker = step_ahead(x, ricker_f, ricker_pardist[i,]), 
                Myers = step_ahead(x, myers_f, myers_pardist[i,]))
  }), id=c("time", "stock"))
}
df_post <- step_ahead_posteriors(x)
library("dplyr")
df_post_means <- summarize(group_by(df_post, time,variable,stock), value = mean(value))
ggplot(df_post_means,aes(time, value)) +
  geom_point(aes(time, stock), size=2) + 
  geom_line(aes(time, value, col=variable, group=variable),lwd=1) + 
#  geom_line(aes(time, value, col=variable, group=interaction(L1,variable))) + 
  facet_wrap(~variable) + 
  scale_colour_manual(values=colorkey) +  
  theme(legend.position="none")

write.csv(df_post_means, "components/data/figure1.csv")
```

@@

```{r figure_2, dependson=c("assemble-models", "plot-options"), fig.cap="The inferred Gaussian process compared to the true process and maximum-likelihood estimated (MLE) process.  We show the expected value for the function $f$ under each model.  Plots show replicates drawn from the posterior distributions in order to convey uncertainty of the estimates.  Note the MLE is a point estimate of parameters and so reflects no uncertainty in the distribution. The training data are also shown as black points.  The GP is conditioned on (0,0), shown as a pseudo-data point.", fig.width=8, fig.height=6, cache=FALSE, dev="CairoJPEG", dev.args=c(quality=100)}
# subset data first, don't care about seeing dynamics for stock >> K
x_grid_short <- x_grid[1:max(which(x_grid<12))]
gp_short <- gp_predict(gp, x_grid_short, burnin=1e4, thin=300)

# sample from subset data
models_posteriors <- 
  melt(lapply(sample(100, 50), 
              function(i){
    sample_gp <- mvrnorm(1, 
                            gp_short$Ef_posterior[,i],         
                            gp_short$Cf_posterior[[i]])
    data.frame(stock = x_grid_short, 
               GP = sample_gp,
               y = sample_gp,
               ymin = sample_gp - 2 * sqrt(gp_short$E_Vf), 
               ymax = sample_gp + 2 * sqrt(gp_short$E_Vf), 
               True = sapply(x_grid_short,f,0, p),  
               MLE = sapply(x_grid_short,f,0, est$p), 
               Allen = sapply(x_grid_short, allen_f, 0, pardist[i,]), 
               Ricker = sapply(x_grid_short, ricker_f, 0, ricker_pardist[i,]), 
               Myers = sapply(x_grid_short, myers_f, 0, myers_pardist[i,]))
             }), 
       id=c("stock", "y", "ymin", "ymax"))
ggplot(models_posteriors) + 
#    geom_ribbon(aes(x=stock, y=y, ymin=ymin, ymax=ymax, group=L1), 
#                  fill = "gray80", 
#                  data=subset(models_posteriors, variable == "GP")) + 
    geom_line(aes(stock, value, col = variable, 
                  group=interaction(L1,variable)), alpha=0.4) + 
    geom_point(data = obs, aes(x,y), size=2) + 
    xlab(expression(X[t])) + ylab(expression(X[t+1])) +
    facet_wrap(~variable) + 
    scale_colour_manual(values=colorkey) +  
    theme(legend.position="none")

@@

We show the  mean inferred population dynamics of each model relative to the
true model used to generate the data in Figure 2, predicting
the relationship between observed population size (x-axis) to the population
size after recruitment the following year.  
In addition to the raw data, the GP is conditioned
on going through the point 0,0 without error.  All parametric models also make this assumption. Conditioning on (0,0) is equivalent to making the assumption that the population is closed, so that once it hits 0 it stays at 0, despite the lack of any data in the observed sequence to justify this.  This assumption illustrates how the GP can capture common-sense biology without having to assume more explicit details about the dynamics at low population numbers that have never been observed. If the population were not closed, one could repeat the entire analysis without this assumption.  Unlike parametric models, the GP corresponds to a
distribution of curves, of which this plot only shows the means. Uncertainty in the parameters of the GP (not shown) further widens the band of possible population sizes.
In Figure S1 (see supplement), we show the performance of the models outside the observed training data.

@@

```{r figure_4, dependson=c("sim"), fig.cap="In the management context, GPDP outperforms approaches based on parametric models. We show 100 replicate simulations of the stock dynamics (Eqn 1) under the policies derived from each of the estimated models, as well as the policy based on the exact underlying model.", fig.width=6, fig.height=6, cache=FALSE, dev="CairoJPEG", dev.args=c(quality=100)}
ggplot(sims_data) + 
  geom_line(aes(time, fishstock, group=interaction(reps,method), color=method), alpha=0.1) +
  scale_colour_manual(values=colorkey, guide = FALSE) +
  xlab("stock size, x(t)") +
  facet_wrap(~method) + 
  guides(legend.position="none", color=FALSE)

@@

Changing the intensity of the stochasticity or the distance between stable and unstable steady states
does not impact the performance of the GP relative to the optimal solution
obtained from the true model and true parameters.  The parametric models are
more sensitive to this difference.  Large values of $\sigma$ relative to the distance between the stable and unstable
point increases the chance of a stochastic transition below the tipping point.  More precisely,
if we let $L$ be the distance between the stable and unstable steady states, then the probability that fluctuations drive the population across the unstable steady state scales as
$\exp\left(\frac{L^2}{\sigma^2}\right)$

@@

```{r figure_6, fig.width=6, fig.height=3, dependson=c("sensitivity-trends"), fig.cap="The effect of increasing noise or decreasing Allee threshold levels on the net present value of the fishery when managed under the GPDP, relative to managing under the true model (with known parameters).  Other than the focal parameter (stochasticity, Allee threshold), other parameters are held fixed as above to illustrate this effect.", cache=FALSE}
plot_sigmas <- ggplot(vary_sigma, aes(noise, value)) + 
  geom_point(size=2) + ylab("Net present value") + 
  xlab("Level of growth stochasticity")
plot_allees <- ggplot(vary_allee, aes(allee, value)) + 
  geom_point(size=2)+ ylab("Net present value") + 
  xlab("Allee threshold stock size")
multiplot(plot_sigmas, plot_allees, cols=2) 
write.csv(vary_sigma, "components/data/figure6a.csv")
write.csv(vary_allee, "components/data/figure6b.csv")
```

@@

Our results illustrate that model-choice approaches can be absolutely
misleading – by providing support to models that cannot capture tipping point
dynamics because they have fewer parameters and the data are far from the tipping point.  That is, when the data 
come from around the stable steady state,
all the parametric models are approximately linear and approximately
identical.  Thus, it is intuitive that all model selection methods choose the simplest model.
In a complex world, the result is suboptimal. But in a world that might
contain tipping points, the result could be disastrous.

Many managers both in fisheries and beyond face a similar situation: they
have not observed the population dynamics at all possible densities.  A
lack of comprehensive data at all population sizes, combined with the
inability to formulate accurate population models for low population sizes
in the absence of data, makes this situation the rule more than the exception.
Relying on parametric models and model choice processes that favor simplicity
ignores this basic reality. For a long time,  Carl Walters [e.g. @Walters1978] has argued that if we began
by fishing any newly exploited population down to very low levels and
then let it recover, we would be much better at estimating population
dynamics and thus predicting the optimal harvest levels.  While certainly true,
this presents a rather risky policy in the face of potential tipping points. 
The GPDP offers a risk-adverse alternative.

@@
